---
title: "Homework 1 - R"
author: "Tim Dennis"
date: "1/8/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Exercise 1 : Making Vectors

Start by making a vector with the numbers 1 through 26.
Multiply the vector by 2, and give the resulting vector
names A through Z (hint: there is a built in vector called `LETTERS`)

### Solution to Exercise 1

```{r}
x <- 1:26
x <- x * 2
names(x) <- LETTERS
```

### Exercise 2 : Matrix

1. Make a matrix containing the numbers 1:50, with 5 columns and 10 rows.
2. Make the matrix above fill your matrix by
row, not by column (its default behaviour). 
(hint: read the documentation for `matrix`!)

### Solution to Exercise 2

Make another matrix, this time containing the numbers 1:50,
with 5 columns and 10 rows.
Did the `matrix` function fill your matrix by column, or by
row, as its default behaviour?
See if you can figure out how to change this.
(hint: read the documentation for `matrix`!)

```{r, eval=FALSE}
x <- matrix(1:50, ncol=5, nrow=10)
x <- matrix(1:50, ncol=5, nrow=10, byrow = TRUE) # to fill by row
```

### Exercise 3 : Data Frame

You can create a new data frame right from within R with the following syntax:
```{r}
df <- data.frame(id = c('a', 'b', 'c'),
                  x = 1:3,
                  y = c(TRUE, TRUE, FALSE),
                  stringsAsFactors = FALSE)
```
Make a data frame that holds the following information for yourself:

- first name
- last name
- lucky number

Then use `rbind` to add an entry for the people sitting beside you.
Finally, use `cbind` to add a column with each person's answer to the question, "Is it time for coffee break?"

### Solution to Exercise 3

```{r}
df <- data.frame(first = c('Grace'),
                 last = c('Hopper'),
                 lucky_number = c(0),
                 stringsAsFactors = FALSE)
df <- rbind(df, list('Marie', 'Curie', 238) )
df <- cbind(df, coffeetime = c(TRUE,TRUE))
```

### Exercise 4 : Lists

Given the following list:

```{r, eval=FALSE}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))
```

Using your knowledge of both list and vector subsetting, extract the **number 2** from xlist.
Hint: the number 2 is contained within the "b" item in the list.

### Solution to exercise 4

```{r, eval=FALSE}
xlist$b[2]
```
```{r, eval=FALSE}
xlist[[2]][2]
```
```{r, eval=FALSE}
xlist[["b"]][2]
```

---

### Exercise 5 : Subsetting gapminder

Fix each of the following common data frame subsetting errors:

1. Extract observations collected for the year 1957

    ```{r, eval=FALSE}
    gapminder[gapminder$year = 1957,]
    ```

2. Extract all columns except 1 through to 4

    ```{r, eval=FALSE}
    gapminder[,-1:4]
    ```

3. Extract the rows where the life expectancy is longer the 80 years

    ```{r, eval=FALSE}
    gapminder[gapminder$lifeExp > 80]
    ```

4. Extract the first row, and the fourth and fifth columns
   (`lifeExp` and `gdpPercap`).

    ```{r, eval=FALSE}
    gapminder[1, 4, 5]
    ```

5. Advanced: extract rows that contain information for the years 2002
    and 2007

    ```{r, eval=FALSE}
    gapminder[gapminder$year == 2002 | 2007,]
    ```

### Solution to Exercise 5

Fix each of the following common data frame subsetting errors:

1. Extract observations collected for the year 1957

```{r, eval=FALSE}
# gapminder[gapminder$year = 1957,]
gapminder[gapminder$year == 1957,]
```

2. Extract all columns except 1 through to 4

```{r, eval=FALSE}
# gapminder[,-1:4]
gapminder[,-c(1:4)]
```

3. Extract the rows where the life expectancy is longer the 80 years

```{r, eval=FALSE}
# gapminder[gapminder$lifeExp > 80]
gapminder[gapminder$lifeExp > 80,]
```

4. Extract the first row, and the fourth and fifth columns (`lifeExp` and `gdpPercap`).

```{r, eval=FALSE}
# gapminder[1, 4, 5]
gapminder[1, c(4, 5)]
```

5. Advanced: extract rows that contain information for the years 2002 and 2007

```{r, eval=FALSE}
# gapminder[gapminder$year == 2002 | 2007,]
gapminder[gapminder$year == 2002 | gapminder$year == 2007,]
gapminder[gapminder$year %in% c(2002, 2007),]
```

---
