---
title: "Homework 1 - R"
author: "Tim Dennis"
date: "1/8/2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Due 1/18 - By Start of Class
## How to submit your homework:
1. Download the r-homework1.R script and open in Rstudio
2. Rename `r-homework1.R` to `yourlastname_sudentID_r-homework1.R`
3. Type the answers in the uncommented lines (those without #). Notice, some lines have gotten you started with names of objects.
4. Save the script file.
5. Attach the file and email to timdennis@ucsd.edu by 1/18 before 6:30pm. NOTE: Please
use your UCSD email address when sending an email.


### Exercise 1 : Making Vectors (7.5 points)

Start by making a vector with the numbers 1 through 26.
Multiply the vector by 2, and give the resulting vector
names A through Z (hint: there is a built in vector called `LETTERS`)

---

### Exercise 2 : Matrix (7.5 points)

1. Make a matrix containing the numbers 1:50, with 5 columns and 10 rows.
2. Make the matrix above fill your matrix by
row, not by column (its default behaviour). 
(hint: read the documentation for `matrix`!)

---

### Exercise 3 : Data Frame (15 points)

You can create a new data frame right from within R with the following syntax:
```{r}
df <- data.frame(id = c('a', 'b', 'c'),
                  x = 1:3,
                  y = c(TRUE, TRUE, FALSE),
                  stringsAsFactors = FALSE)
```
Make a data frame that holds the following information for yourself:

- first name
- last name
- lucky number

Then use `rbind` to add an entry for the people sitting beside you.
Finally, use `cbind` to add a column with each person's answer to the question, "Is it time for coffee break?"

---

### Exercise 4 : Lists (5 points)

Given the following list:

```{r, eval=FALSE}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))
```

Using your knowledge of both list and vector subsetting, extract the **number 2** from xlist.
Hint: the number 2 is contained within the "b" item in the list.

---

### Exercise 5 : Subsetting gapminder (15 points)

Fix each of the following common data frame subsetting errors:

1. Extract observations collected for the year 1957

    ```{r, eval=FALSE}
    gapminder[gapminder$year = 1957,]
    ```

2. Extract all columns except 1 through to 4

    ```{r, eval=FALSE}
    gapminder[,-1:4]
    ```

3. Extract the rows where the life expectancy is longer the 80 years

    ```{r, eval=FALSE}
    gapminder[gapminder$lifeExp > 80]
    ```

4. Extract the first row, and the fourth and fifth columns
   (`lifeExp` and `gdpPercap`).

    ```{r, eval=FALSE}
    gapminder[1, 4, 5]
    ```

5. Advanced: extract rows that contain information for the years 2002
    and 2007

    ```{r, eval=FALSE}
    gapminder[gapminder$year == 2002 | 2007,]
    ```
