---
title: "Plotting with GGPLOT2"
author: "Tim Dennis & Reid Otsuji"
date: "January 8, 2016"
output: 
  html_document: 
    toc: yes
---

**start hopefully 3:25**


```{r, include=FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/08-plot-ggplot2-")
# Silently load in the data so the rest of the lesson works
#gapminder <- read.csv("data/gapminder-FiveYearData.csv", header=TRUE)
```
```{r}
#install.packages('ggplot2')
#library(ggplot2)
```


GOALS: Students should be able to use ggplot2 to generate publication quality graphics and understand and use the basics of the grammar of graphics.

* GGPLOT2 developed by Hadley Wickham based on a *grammar-of-graphics*
* Grammar of graphics consists of a **dataset**, **coordinate system**, and **geoms** -- the visual representation of data points.
* Think about figure in layers:  like you would in Photoshop, illustrator or Inkscape.

##DataViz

* plotting our data is one of the best was to quickly explore it and the various relationships between variables
* 3 main plotting systems in R: the base [plotting system](http://www.statmethods.net/graphs/), the [lattice package](http://www.statmethods.net/advgraphs/trellis.html), and [ggplot2](http://www.statmethods.net/advgraphs/ggplot2.html)
*`ggplot2` is built on the grammar-of-graphics:
   * idea that any plot can be expressed from the same set of components: a **data** set, a **coodinate system** and a set of **geoms**-- the visual representation of data points
* **Key** to understanding `ggplot2` is thinking about a figure in layers -- think of ArcGIS or programs like Photoshop

## Terminology:

* **ggplot** - The main function where you specify the dataset and variables to plot
* **geoms** - geometric objects
  * `geom_point()`, `geom bar()`, `geom density()`, `geom line()`, `geom area()`
* **aes** - aesthetics 
  * shape, transparency (alpha), color, fill, line-type.
* **scales** Define how your data will be plotted
  * continuous, discrete, log

## First Plots with GGPLOT

```{r lifeExp-vs-gdpPercap-scatter, message=FALSE}
#gapminder <- read.csv("https://goo.gl/BtBnPg", header = T)
gapminder <- read.csv('data/gapminder-FiveYearData.csv', header=T)
```

Let's start off with an example:

```{r gap-scatter}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

NOTE:

* First we call the `ggplot` function -any arguments we provide the `ggplot` function are considered **global** options: they apply to all layers
* We passed two arguments to `ggplot`:
  * `data`
  * an `aes` function - which tells ggplot how variables map to aesthetic properties
  * `x` & `y` locations
  * notice we didn't have to define the data and column

Alone the `ggplot` call isn't enough to render the plot.

```{r no_geom, eval=FALSE}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap))
## If run, would produce an error.
```

* Need to tell **ggplot** how we want to present variables by specifying a geom layer. * In the above example we used `geom_point` to create a scatter plot.

```{r lifeExp-vs-gdpPercap-scatter2}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```

### Challenge 1

<http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2#challenge-1>

### Challenge 2

<http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2#challenge-2>

## Layers

* Using scatter plot not the best way to visualize change over time. 
* Let's use line plot.

```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line()
```

* used `geom_line` instead of `geom_point`
* added a **by** *aesthetic* to get a line per country and color by continent


* visualize both lines and points on the plot?

```{r lifeExp-line-point}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line() + geom_point()
```

* important to note this is layered: so points have been drawn *on top of* the lines.
* example of this

```{r lifeExp-layer-example-1}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country)) +
  geom_line(aes(color=continent)) + geom_point()
```

* in the above the *aesthetic* mapping of **color** has been moved from the
global plot options in `ggplot` to the `geom_line` layer so it no longer applies to the points

### Challenge 3

<http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2#challenge-3>

> > ## Solution to challenge 3
> >
> > Switch the order of the point and line layers from the previous example. What
> > happened?
> >
> > ```{r ch3-sol}
> > ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country)) +
> >  geom_point() + geom_line(aes(color=continent))
> > ```
> >
> > The lines now get drawn over the points!
> >


## Transformations and statistics

* easy to overlay statistical models over the data
* let's use our first example

```{r lifeExp-vs-gdpPercap-scatter3, message=FALSE}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, color=continent)) +
  geom_point()
```

* hard to see relationships b/t points because of strong outliers in GDP/cap
* We can change the scale of units on the `y` axis using the *scale* functions
* These control the mapping between the data values and visual values of an aesthetic.
* We'll also use the `alpha` function, which is helpful when you have a large amount of data which is v. clustered

```{r axis-scale}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point(alpha=0.5) + scale_y_log10()
```

* `log10` function applied a transformation to the values of the gdpPercap column before rendering them on the plot
   * each multiple of 10 now only corresponds to an increase in 1 on the transformed scale, 
   * e.g. a GDP per capita of **1,000 is now 3** on the y axis, a value of 10,000 corresponds to 4 on the y axis and so on
* This makes it easier to visualize the spread of data on the y-axis.

* We can fit a simple relationship to the data by adding another layer, `geom_smooth`:

```{r lm-fit}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10() + geom_smooth(method="lm")
```

*  make the line thicker by *setting* the **size** aesthetic in the `geom_smooth` layer:

```{r lm-fit2}
pwd <- ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10() + geom_smooth(method="lm", size=1.5)
```

* two ways an *aesthetic* can be specified:
  1. Here we *set* the **size** aesthetic by passing it as an argument to `geom_smooth`.
  2. use the `aes` function to define a *mapping* between data variables and their visual representation.

### Challenge 4a

<http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2#challenge-3#challenge-4a>

### Challenge 4b

<http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2#challenge-3#challenge-4b>

## Multi-panel figures: FACEting

* previously we visualized the change in life expectancyover time across all countries in one plot
* we can split this out over multiple panels by adding a layer of **facet** panels
* to limit the number of panels let's work only on countries beginning in the letter "A" or "Z"

```{r startswa-z}
starts.with <- substr(gapminder$country, start = 1, stop = 1)
az.countries <- gapminder[starts.with %in% c("A", "Z"), ]
```

Talk thru code: 
* We'll start by subsetting the data using the `substr` function topull out a part of a character string; 
* in this case, the letters that occur in positions `start` through `stop`, inclusive, of the `gapminder$country` vector. 
* The operator `%in%` allows us to make multiple comparisons rather than write out long subsetting conditions (in this case, `starts.with %in% c("A", "Z")` is equivalent to `starts.with == "A" | starts.with == "Z"`)

```{r facet}
ggplot(data = az.countries, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country)
```

* `facet_wrap` layer took a "formula" as its argument, denoted by the tilde
(~).
* tells R to draw a panel for each unique value in the country column
of the gapminder dataset.

## Modifying text

* would like to add text to elements in the graph
* do this by adding a few more layers:
   * **theme** layer controls axis text & text size
   * **scales** layer to change legend title

```{r theme}
ggplot(data = az.countries, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country) +
  xlab("Year") + ylab("Life expectancy") + ggtitle("Figure 1") +
  scale_colour_discrete(name="Continent") +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())
```

<http://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2#challenge-5>

This is a taste of what you can do with `ggplot2`. RStudio provides a
really useful [cheat sheet][cheat] of the different layers available, and more
extensive documentation is available on the [ggplot2 website][ggplot-doc].
Finally, if you have no idea how to change something, a quick Google search will
usually send you to a relevant question and answer on Stack Overflow with reusable
code to modify!

## How to save your plots

```{r eval=FALSE}
ggsave('~/path/to/figure/filename.png')
ggsave(plot1, file = "~/path/to/figure/filename.png")
```

```{r eval=FALSE}
ggsave(file = "/path/to/figure/filename.png", width = 6,
height =4)
```

```{r eval=FALSE}
ggsave(file = "/path/to/figure/filename.eps")
ggsave(file = "/path/to/figure/filename.jpg")
ggsave(file = "/path/to/figure/filename.pdf")
```

## Resources:

This is just a taste of what you can do with `ggplot2`. RStudio provides a
really useful [cheat sheet][cheat] of the different layers available, and more
extensive documentation is available on the [ggplot2 website][ggplot-doc].
Finally, if you have no idea how to change something, a quick Google search will
usually send you to a relevant question and answer on Stack Overflow with reusable
code to modify!

[cheat]: http://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf
[ggplot-doc]: http://docs.ggplot2.org/current/
