---
title: "r-homework1-edit"
author: "Reid Otsuji"
date: "1/11/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Due 1/17 - By Start of Class 6:30pm
## How to submit your homework:
1. Download the [r-homework1.R](https://raw.githubusercontent.com/ucsdlib/win2018-gps-r/gh-pages/Homework/r-homework1.R) (save as from your browser) script and open in Rstudio
2. Rename `r-homework1.R` to `yourlastname_sudentID_r-homework1.R` (save as in RStudio)
3. Type the answers in the uncommented lines (those without #). Notice, for some lines, we have gotten you started with names of objects.
4. Save the script file.
5. Upload the file to tritonEd (preferred). By 1/17 before 6:30pm. NOTE: Please use your UCSD email address when sending an email.  



### Exercise 1 : Making Vectors (7.5 points)

Start by making a vector named 'myvector' with the numbers 1 through 26.
Create another vector named, 'myvectimestwo' by multiply the vector by 2, and give the resulting vector names A through Z (hint: there is a built in vector called `LETTERS`)

---
### Exercise  : Help
Start by checking the preloaded datasets by typing:
```{r eval=FALSE}
data() 
```




Find the `mtcars` data set. Write the command to find the `help` information about the mtcars dataset.
?mtcars




what does the dataset represent?
---

### Exercise: Factors
Load the small cats dataset we created in class (feline-data.csv).  If you don't have this data set, see the intro lesson.
```{r eval=FALSE}
cats <- read.csv(file="feline-data.csv")
```

Try using `?read.csv` to figure out how to keep text columns (e.g. coats) as character vectors instead of factors; then write a command or two to show that the factor in `cats` is actually a `character vector` when loaded in this way.


hint: start with:
```{r eval=FALSE}
cats <- read.csv()
str()
```


solution:
cats <- read.csv(file="feline-data.csv", stringAsFactors=FALSE)
str(cats$coat)


---
### Exercise: Data Types and Structures

dataTypes <- c('double', 'complex', 'integer', 'character', 'logical')
dataStructures <- c('data.frame', 'vector', 'factor', 'list', 'matrix')
answer <- list(dataTypes, dataStructures)

---

### Exercise  : Structure 
1. Use the function `data()` to display the list of Locate the builtin `iris` dataset. 
```{r eval=FALSE}
data()
```

2. Write function used to investigate the structure of the dataset `iris`.

solution:
```{r eval=FALSE}
str(iris)
```


3. Copy the data.frame output from the command output?
# copy and past output here:
```{r eval=FALSE}
'data.frame':	150 obs. of  5 variables:
 $ Sepal.Length: num  5.1 4.9 4.7 4.6 5 5.4 4.6 5 4.4 4.9 ...
 $ Sepal.Width : num  3.5 3 3.2 3.1 3.6 3.9 3.4 3.4 2.9 3.1 ...
 $ Petal.Length: num  1.4 1.4 1.3 1.5 1.4 1.7 1.4 1.5 1.4 1.5 ...
 $ Petal.Width : num  0.2 0.2 0.2 0.2 0.2 0.4 0.3 0.2 0.2 0.1 ...
 $ Species     : Factor w/ 3 levels "setosa","versicolor",..: 1 1 1 1 1 1 1 1 1 1 ...
```

4. How many objects(obs) and variables are in the dataset? 

solution:
150 obs, 5 variables

5. list the commands you would use to find the levels attribute for the `species` varaiable. 
(Hint: start with assigning the iris database to a variable `df <- iris` . Use `levels()` function.)
```{r eval=FALSE}
df <- iris
levels() #complete this line.
```


solution:
df <- iris
levels(df, species)
---

### Exercise 2 : Matrix (7.5 points)

1. Make a matrix called "mymatrix" containing the numbers 1:50, with 5 columns and 10 rows. Print the matrix out.
2. Make the "mymatrix" matrix above fill by row, not by column (its default behaviour). Print out the matrix.
(hint: read the documentation for `matrix`!)
 
---

<!--
### Exercise 3 : Data Frame (15 points)

You can create a new data frame right from within R with the following syntax:
```{r}
df <- data.frame(id = c('a', 'b', 'c'),
                  x = 1:3,
                  y = c(TRUE, TRUE, FALSE),
                  stringsAsFactors = FALSE)
```
Make a data frame that holds the following information for yourself:

- first name
- last name
- lucky number

Column names should be first_name, last_name, & lucky_number
Then use `rbind` to add an entry for someone else in the class or someone you know.
Finally, use `cbind` to add a column named 'coffee' with each person's answer to the question, "Is it time for coffee break?"

---

-->

### Exercise 4 : Lists (5 points)

Given the following list:

```{r, eval=FALSE}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))
```

Using your knowledge of both list and vector subsetting, extract the **number 2** from xlist.
Hint: the number 2 is contained within the "b" item in the list.  Look up how to subset lists
from the lecture notes. 

---

