---
title: "06_subsetting_lesson_challenges"
output: html_document
---

## Challenge 1

Given the following code:

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```

Come up with at least 3 different commands that will produce the following output:

```{r, echo=FALSE}
x[2:4]
```

After you find 3 different commands, compare notes with your neighbour. Did you have different strategies?

---

## Challenge 2

Run the following code to define vector `x` as above:

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```

Given this vector `x`, what would you expect the following to do?

```{r, eval=FALSE}
x[-which(names(x) == "g")]
```

Try out this command and see what you get. Did this match your expectation?

Why did we get this result? (Tip: test out each part of the command on it's own - this is a useful debugging strategy)

Which of the following are true:

* A) if there are no `TRUE` values passed to `which`, an empty vector is returned
* B) if there are no `TRUE` values passed to `which`, an error message is shown
* C) `integer()` is an empty vector
* D) making an empty vector negative produces an "everything" vector
* E) `x[]` gives the same result as `x[integer()]`


---

## Challenge 3

Given the following code:

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```

Write a subsetting command to return the values in x that are greater than 4 and less than 7.


---

## Challenge 4

Given the following code:

```{r}
m <- matrix(1:18, nrow=3, ncol=6)
print(m)
```

1. Which of the following commands will extract the values 11 and 14?

A. `m[2,4,2,5]`

B. `m[2:5]`

C. `m[4:5,2]`

D. `m[2,c(4,5)]`



---


## Challenge 5
Given the following list:

```{r, eval=FALSE}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))
```

Using your knowledge of both list and vector subsetting, extract the number 2 from xlist.
Hint: the number 2 is contained within the "b" item in the list.


---

## Challenge 6

Given a linear model:

```{r, eval=FALSE}
mod <- aov(pop ~ lifeExp, data=gapminder)
```

Extract the residual degrees of freedom (hint: `attributes()` will help you)



---

## Challenge 7

Fix each of the following common data frame subsetting errors:

1. Extract observations collected for the year 1957

```{r, eval=FALSE}
   gapminder[gapminder$year = 1957,]
```

2. Extract all columns except 1 through to 4

```{r, eval=FALSE}
   gapminder[,-1:4]
```

3. Extract the rows where the life expectancy is longer the 80 years

```{r, eval=FALSE}
   gapminder[gapminder$lifeExp > 80]
```

4. Extract the first row, and the fourth and fifth columns
  (`lifeExp` and `gdpPercap`).

```{r, eval=FALSE}
   gapminder[1, 4, 5]
```

5. Advanced: extract rows that contain information for the years 2002 and 2007

```{r, eval=FALSE}
   gapminder[gapminder$year == 2002 | 2007,]
```


 
---
 
## Challenge 8

1. Why does `gapminder[1:20]` return an error? How does it differ from `gapminder[1:20, ]`?


2. Create a new `data.frame` called `gapminder_small` that only contains rows 1 through 9
and 19 through 23. You can do this in one or two steps.


