---
title: "GGPLOT2 Lesson Challenges"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Challenge 1

Modify the example so that the figure visualise how life expectancy has changed over time:

Hint: the gapminder dataset has a column called “year”, which should appear on the x-axis.

```{r, eval=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) + geom_point()
```

---

## Challenge 2

In the previous examples and challenge we’ve used the aes function to tell the scatterplot geom about the x and y locations of each point. Another aesthetic property we can modify is the point color. Modify the code from the previous challenge to color the points by the “continent” column. What trends do you see in the data? Are they what you expected?

---

## Challenge 3

Switch the order of the point and line layers from the previous example. What happened?

previous example:
http://swcarpentry.github.io/r-novice-gapminder/fig/rmd-08-lifeExp-layer-example-1-1.png


---

## Challenge 4a
Modify the color and size of the points on the point layer in the previous example.

```{r, eval=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + scale_x_log10() + geom_smooth(method="lm", size=1.5)
```

Hint: do not use the aes function.



---

## Challenge 4b
Modify your solution to Challenge 4a so that the points are now a different shape and are colored by continent with new trendlines. 

Hint: The color argument can be used inside the aesthetic.


```{r, eval=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
 geom_point(size=3, color="orange") + scale_x_log10() +
 geom_smooth(method="lm", size=1.5)
```

---

## Challenge 5

Create a density plot of GDP per capita, filled by continent.

Advanced: - Transform the x axis to better visualise the data spread. - Add a facet layer to panel the density plots by year.

