---
title: "Intro to R"
author: "Reid Otsuji adapted from Tim Dennis"
date: "January 5, 2018"
output: html_document
---

**shoot for 1hr**

## SETUP (5 min)

* We're going to learn R for data analysis!
* real goal not to teach R indepth, but teach basic concepts that you can use in any language
* we’re going to teach you some of the fundamentals of the R language as well as some best practices for organizing code for scientific projects that will make your life easier

* why R and RStudio?
	* we have to use something
	* it's free, well-documented, and runs on most machines
	* active community, especially among scientists
	* lots of code already written (and available for reuse!) for many different types of analyses
	
## SETUP DATA
* download data and put in SWC workshop folder: https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv

## INTRO TO RSTUDIO (10 min)

* RStudio is a free, handy interface to work with R (some folks may have used the regular R console already)
	* R works the same, regardless of whether you're using the R console, R GUI, or RStudio
	
##### BASIC LAYOUT: RStudio
* When you open RStudio, you will be greeted by three panels:
    * The interactive R console (entire left)
    * Workspace/History (tabbed in upper right)
    * files/Plots/Packages/Help (tabbed in lower right)
    
* Walk through each panel on screen

##### KEYBOARD SHORTCUTS
* Shortcuts: 
`Tools>Keyboard shortcut help`

Alt+- inserts `<-` at cursor
Control+Shift+m inserts `%>%` at cursor (we'll get into thisfor day 2)
Control+Enter = Run current line/selection
ctrl+1 = Move cursor to source
ctrl+2 = Move cursor to console

## Intro to R (15)

* R as calculator - simple use 
	* much time spent in console working out code 
	* console `>` with blinking cursor, much like command line
	* "read, evaluate, print, loop" REPL - many languages adopt this paradigm (bash, stata, python)

```{r eval=FALSE}
1 + 100
1 + 
```

* if you hit return and R shows a `+` instead of ~>`, waiting for you to comlete command, ESC or control+c will escape

R uses same order of opterations to lowest precedence:

1. Parentheses: `( )`
2. Exponents: `^` or `**`
3. Divide: `/`
4. Multiply: `*`
5. Add: +
6. Subtract: -

`3+5*2` v. `(3+5) * 2`

* really small or large numbers get scientific notation
* `2/10000` #2e-04 shorthand for 10^XX 
* So 2e-4 is shorthand for 2 * 10^(-4).
* `5e3` #notice the lack of minus here

## Math functions - many built in functions

```{r}
sin(1) #trig functions
log(1) # natural log
log10(10) #base-10 log
```

* notice the use of the `#` after, any idea what this does? this doesn't get evaluated b/c it's a comment, use this to document or leave notes for yourself, e.g. #TODO fix code
* Dont worry about remembering functions, use google 
* use RSudio's **autocompletion** feature if you can remember beginning of function
* Typing `?` before function brings up help page in Rstudio

## Comparing things

```{r}
1 == 1 # equality (note two equals signs, read as "is equal to")
1 != 2  # inequality (read as "is not equal to")
1 <  2  # less than
1 <= 1  # less than or equal to
1 > 0  # greater than
1 >= -9 # greater than or equal to
# Tip: dont' use == to compare numbers unless integers, computers represent decimals with a certain degree of precision
# check out ?all.equal for comparing things involving doubles

0.1+0.05==0.15
all.equal(0.1+0.05, 0.15)
```

* Tip: dont' use == to compare numbers unless integers, computers represent decimals with a certain degree of precision
* ?all.equal
* `0.1+0.05==0.15`
* `all.equal(0.1+0.05, 0.15)`

## Variable & assignments

* Store values in vars using assignment operator <-, like `x <- 1/40`
* Look in Environment tab in Rstudio
* our var can be used in place of a number in calculations `log(x)`
* Variables can be reassigned `x<-100`
* Assignement values can contain variable being assigned to
* `x <- x + 1 #notice how RStudio updates its description of x on the top right tab`
* right hand side can be any valid R expression & is evaluated prior assigment
* Variable names can contain letters, numbers, underscores and periods. 
* They cannot start with a number nor contain spaces at all.
* Can use = for assignment but less common among R users

## Challenge 1
<https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro/#challenge-1>

## Vectorization

* R is vectorized, meaning that variables and functions can have vectors as values

```{r}
1:5
2^(1:5)
x <- 1:5
2^x
```

## Challenge 2 & 3
<https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro/challenge-2>
<https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro/challenge-3>

## Managing your enviorment

* usefule commands for working with R session 
* `ls()` shows R ojects
* `ls` plain will print out code for that function (or any R function)
* Remove objects use `rm(x)`
* If you have a lot of objects and want to delete all, use rm(list=ls())
* In this case we are using ls() function inside another function that takes a list argument, so we are listing all objects and then deleting them with rm()
* arguments need the =, not <- `rm(list <- ls())` causes errors

<https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro/challenge-4>

## R packages

* We can write our own packages or obtain package written by others.
* Over 7,000 packages on CRAN (the comprehensive R archive network)
* In RStudio console run: 

```{r eval=FALSE}
installed.packages() #list packages
install.packages("packagename1", "packagename2") #install one or many packages
update.pakcages()
remove.packages("packagename")
library(packagename) #make package available 

```

<https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro/challenge-5>

#### Project management with RStudio (10 min)

* most ppl tend to organize projects like so
* <http://swcarpentry.github.io/r-novice-gapminder/fig/bad_layout.png>

> ## Creating a project in RStudio
> We’re going to create a new project in RStudio:
>
> 1. Click the “File” menu button, then “New Project”.
> 2. Click “New Directory”.
> 3. Click “Empty Project”.
> 4. Type in the name of the directory to store your project, e.g. “swc_ucla”.
> 5. If available, select the checkbox for “Create a git repository." (We'll come back to this tomorrow) 
> 6. Click the “Create Project” button.

<https://swcarpentry.github.io/r-novice-gapminder/02-project-intro/#challenge-1>

## Seeking help (5min)

To be able to access R help files for functions and operators. 

* R & every package provides help files for function. 
* to search: `?function_name` or `help(function_name)` 
* special operators:  `?"+"`
* packages can also have "vignettes" -- tutorials 
* use vignette() to see all
* vignette('vignette_name')
* `??function_name` will do fuzzy search for function help (if you don't know the exact name)
* don't know where to begin to find a package you need, look at CRAN taask views http://cran.at.r-project.org/web/views/ (cat based )

** hope to be at 2:00pm**
