---
title: "GGPLOT2 Lesson Challenges"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Challenge 1

Modify the example figure so that the new figure visualise how life expectancy has changed over time:

Example figure: http://swcarpentry.github.io/r-novice-gapminder/fig/rmd-08-lifeExp-vs-gdpPercap-scatter2-1.png

Hint: the gapminder dataset has a column called “year”, which should appear on the x-axis.

```{r, eval=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) + geom_point()
```

---

## Challenge 2

In the previous examples and challenge in the lesson we’ve used the **aes** function to tell the scatterplot geom about the **x** and **y** locations of each point. Another aesthetic property we can modify is the **point color**. Modify the code from the previous challenge to color the points by the “continent” column. What trends do you see in the data? Are they what you expected?

---

## Challenge 3

Switch the order of the point and line layers from the previous example. What happened?

previous example:
http://swcarpentry.github.io/r-novice-gapminder/fig/rmd-08-lifeExp-layer-example-1-1.png


---

## Challenge 4a
Modify the color and size of the points on the point layer in the previous example.

Previous example: http://swcarpentry.github.io/r-novice-gapminder/fig/rmd-08-lm-fit2-1.png

```{r, eval=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + scale_x_log10() + geom_smooth(method="lm", size=1.5)
```

Hint: do not use the aes function.



---

## Challenge 4b
Modify your solution to Challenge 4a so that the points are now a different shape and are colored by continent with new trendlines. 

Solution 4a Plot: http://swcarpentry.github.io/r-novice-gapminder/fig/rmd-08-ch4a-sol-1.png

Hint: The color argument can be used inside the aesthetic.


```{r, eval=FALSE}
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
 geom_point(size=3, color="orange") + scale_x_log10() +
 geom_smooth(method="lm", size=1.5)
```

---

## Challenge 5

Create a density plot of GDP per capita, filled by continent.

Advanced: - Transform the x axis to better visualise the data spread. - Add a facet layer to panel the density plots by year.

