---
title: "2017 R Lesson Challenges"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

## Lesson 1: Intro to R Challenges

### Challenge 1 -- Added to quiz 

Which of the following are valid R variable names?
```{r, eval=FALSE}
min_height
max.height
 _age
.mass
MaxLength
min-length
 2widths
celsius2kelvin
```

### Solution to challenge 1

The following can be used as R variables:
```{r, eval=FALSE}
min_height
max.height
MaxLength
celsius2kelvin
```

The following creates a hidden variable:
```{r, eval=FALSE}
.mass
```

The following will not be able to be used to create a variable
```{r, eval=FALSE}
_age
min-length
2widths
```
---

### Challenge 2 - already in quiz (maybe rewrite if needed)

TODO: poss. rewrite if needed (change var names, etc.)

What will be the value of each  variable  after each 
statement in the following program?


```{r, eval=FALSE}
mass <- 47.5
age <- 122
mass <- mass * 2.3
age <- age - 20
```

### Solution to challenge 2

```{r ch2pt1-sol}
mass <- 47.5
```
This will give a value of `r mass` for the variable mass

```{r ch2pt2-sol}
age <- 122
```
This will give a value of `r age` for the variable age

```{r ch2pt3-sol}
mass <- mass * 2.3
```
This will multiply the existing value of `r mass/2.3` by 2.3 to give a new value of
`r mass` to the variable mass.

```{r ch2pt4-sol}
age <- age - 20
```
This will subtract 20 from the existing value of `r age + 20 ` to give a new value
of `r age` to the variable age.

---

### Challenge 3 - in quiz
Run the code from the previous challenge, and write a command to
compare mass to age. Is mass larger than age?


### Solution to challenge 3

One way of answering this question in R is to use the `>` to set up the following:
```{r, eval=TRUE}
mass > age
```

This should yield a boolean value of TRUE since `r mass` is greater than `r age`.

---

### Challenge 4  inclass

Clean up your working environment by deleting the mass and age
variables.

### Solution to challenge 4

We can use the `rm` command to accomplish this task
```{r ch4-sol}
rm(age, mass)
```
---

### Challenge 5 - already in quiz
Install the following packages: `ggplot2`, `plyr`, `gapminder`

### Solution to challenge 5

We can use the `install.packages()` command to install the required packages.
```{r ch5-sol, eval=FALSE}
install.packages("ggplot2")
install.packages("plyr")
install.packages("gapminder")
```

---

## Lesson 2: Project Management with RStudio 

### Challenge 1 - good for class! 

Download the gapminder data from [here](https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv).

1. Download the file (CTRL + S, right mouse click -> "Save as", or File -> "Save page as")
2. Make sure it's saved under the name `gapminder-FiveYearData.csv`
3. Save the file in the `data/` folder within your project.

We will load and inspect these data later.

---


### Challenge 2 - good for class 

It is useful to get some general idea about the dataset, directly from the
command line, before loading it into R. Understanding the dataset better
will come handy when making decisions on how to load it in R. Use command-line
shell to answer the following questions:

1. What is the size of the file?
2. How many rows of data does it contain?
3. What are the data types of values stored in this file

### Solution to Challenge 2

By running these commands in the shell:
```{r, engine='bash' }
#ls -lh ~/data/gapminder-FiveYearData.csv
ls -lh ~/workshops/lecture-notes/gps-skills-2017/r-intro/data/gapminder-FiveYearData.csv
```


The file size is 80K.
```{r, engine='bash' }
#wc -l ~/data/gapminder-FiveYearData.csv
wc -l ~/workshops/lecture-notes/gps-skills-2017/r-intro/data/gapminder-FiveYearData.csv
```


There are 1705 lines and the data looks like:
```{r, engine='bash' }
#head ~/data/gapminder-FiveYearData.csv
head ~/workshops/lecture-notes/gps-skills-2017/r-intro/data/gapminder-FiveYearData.csv
```

---

### Challenge 3 - in class 

1. Create a directory within your project called `graphs`.
2. Modify the `.gitignore` file to contain `graphs/`
so that this disposable output isn't versioned.

Add the newly created folders to version control using
the git interface.

### Solution to Challenge 3-in class

This can be done with the command line:
```
$ mkdir graphs
$ echo "graphs/" >> .gitignore
```

---

## Lesson 3: Seeking Help 

### Challenge 1 - inclass

Look at the help for the `c` function. What kind of vector do you
expect you will create if you evaluate the following:
```{r, eval=FALSE}
c(1, 2, 3)
c('d', 'e', 'f')
c(1, 2, 'f')`
```

### Solution to Challenge 1

The `c()` function creates a vector, in which all elements are the
same type. In the first case, the elements are numeric, in the
second, they are characters, and in the third they are characters:
the numeric values "coerced" to be characters.

---

### Challenge 2 - in the quiz 

Look at the help for the `paste` function. You'll need to use this later.
What is the difference between the `sep` and `collapse` arguments?

### Solution to Challenge 2

Look at the help for the `paste` function. You'll need to use this later.

```{r, eval=FALSE}
help("paste")
?paste
```

---

### Challenge 3 - in class

Use help to find a function (and its associated parameters) that you could
use to load data from a csv file in which columns are delimited with "\t"
(tab) and the decimal point is a "." (period). This check for decimal
separator is important, especially if you are working with international
colleagues, because different countries have different conventions for the
decimal point (i.e. comma vs period).
hint: use `??csv` to lookup csv related functions.

---

## Lesson 4: Data Structures - Part 1 

### Challenge 1 - inclass or homework

Start by making a vector with the numbers 1 through 26.
Multiply the vector by 2, and give the resulting vector
names A through Z (hint: there is a built in vector called `LETTERS`)

### Solution to Challenge 1

```{r}
x <- 1:26
x <- x * 2
names(x) <- LETTERS
```

---


### Challenge 2 - good for class

Is there a factor in our `cats` data.frame? what is its name?
Try using `?read.csv` to figure out how to keep text columns as character
vectors instead of factors; then write a command or two to show that the factor
in `cats` is actually a character vector when loaded in this way.

### Solution to Challenge 2

One solution is use the argument `stringAsFactors`:

```{r, eval = FALSE}
cats <- read.csv(file="data/feline-data.csv", stringsAsFactors=FALSE)
str(cats$coat)
```

Another solution is use the argument `colClasses`
that allow finer control.

```{r, eval = FALSE}
cats <- read.csv(file="data/feline-data.csv", colClasses=c(NA, NA, "character"))
str(cats$coat)
```

---


### Challenge 3 - inclass

There are several subtly different ways to call variables, observations and
elements from data.frames:

- `cats[1]`
- `cats[[1]]`
- `cats$coat`
- `cats["coat"]`
- `cats[1, 1]`
- `cats[, 1]`
- `cats[1, ]`

Try out these examples and explain what is returned by each one.

*Hint:* Use the function `typeof()` to examine what is returned in each case.

## Solution to Challenge 3
```{r, eval=FALSE, echo=TRUE}
cats[1]
```
We can think of a data frame as a list of vectors. The single brace `[1]`
returns the first slice of the list, as another list. In this case it is the
first column of the data frame.
```{r, eval=FALSE, echo=TRUE}
cats[[1]]
```
The double brace `[[1]]` returns the contents of the list item. In this case
it is the contents of the first column, a _vector_ of type _factor_.
```{r, eval=FALSE, echo=TRUE}
cats$coat
```
This example uses the `$` character to address items by name. _coat_ is the
first column of the data frame, again a _vector_ of type _factor_.
```{r, eval=FALSE, echo=TRUE}
cats["coat"]
```
Here we are using a single brace `["coat"]` replacing the index number with
the column name. Like example 1, the returned object is a _list_.
```{r, eval=FALSE, echo=TRUE}
cats[1, 1]
```
This example uses a single brace, but this time we provide row and column
coordinates. The returned object is the value in row 1, column 1. The object
is an _integer_ but because it is part of a _vector_ of type _factor_, R
displays the label "calico" associated with the integer value.
```{r, eval=FALSE, echo=TRUE}
cats[, 1]
```
Like the previous example we use single braces and provide row and column
coordinates. The row coordinate is not specified, R interprets this missing
value as all the elements in this _column_ _vector_.
```{r, eval=FALSE, echo=TRUE}
cats[1, ]
```
Again we use the single brace with row and column coordinates. The column
coordinate is not specified. The return value is a _list_ containing all the
values in the first row.

---

### Challenge 4 -in class

What do you think will be the result of
`length(matrix_example)`?
Try it.
Were you right? Why / why not?

### Solution to Challenge 4

What do you think will be the result of
`length(matrix_example)`?

```{r}
matrix_example <- matrix(0, ncol=6, nrow=3)
length(matrix_example)
```

Because a matrix is a vector with added dimension attributes, `length`
gives you the total number of elements in the matrix.


---

### Challenge 5 - homework

Make another matrix, this time containing the numbers 1:50,
with 5 columns and 10 rows.
Did the `matrix` function fill your matrix by column, or by
row, as its default behaviour?
See if you can figure out how to change this.
(hint: read the documentation for `matrix`!)

### Solution to Challenge 5

Make another matrix, this time containing the numbers 1:50,
with 5 columns and 10 rows.
Did the `matrix` function fill your matrix by column, or by
row, as its default behaviour?
See if you can figure out how to change this.
(hint: read the documentation for `matrix`!)

```{r, eval=FALSE}
x <- matrix(1:50, ncol=5, nrow=10)
x <- matrix(1:50, ncol=5, nrow=10, byrow = TRUE) # to fill by row
```

---

### Challenge 6 - inclass
Create a list of length two containing a character vector for each of the sections in this part of the workshop:

- Data types
- Data structures

Populate each character vector with the names of the data types and data
structures we've seen so far.

### Solution to Challenge 6

```{r}
dataTypes <- c('double', 'complex', 'integer', 'character', 'logical')
dataStructures <- c('data.frame', 'vector', 'factor', 'list', 'matrix')
answer <- list(dataTypes, dataStructures)
```
Note: it's nice to make a list in big writing on the board or taped to the wall
listing all of these types and structures - leave it up for the rest of the workshop
to remind people of the importance of these basics.

---

### Challenge 7 - class

Consider the R output of the matrix below:
```{r, eval=FALSE}
matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
```
What was the correct command used to write this matrix? Examine
each command and try to figure out the correct one before typing them.
Think about what matrices the other commands will produce.

1. `matrix(c(4, 1, 9, 5, 10, 7), nrow = 3)`
2. `matrix(c(4, 9, 10, 1, 5, 7), ncol = 2, byrow = TRUE)`
3. `matrix(c(4, 9, 10, 1, 5, 7), nrow = 2)`
4. `matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)`

### Solution to Challenge 7

Consider the R output of the matrix below:
```{r, eval=FALSE}
matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
```
What was the correct command used to write this matrix? Examine
each command and try to figure out the correct one before typing them.
Think about what matrices the other commands will produce.
```{r, eval=FALSE }
matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)
```

---

## Lesson 5: Exploring Data Frames - Data Structures - Part 2 

---


### Challenge 1 - inclass

You can create a new data frame right from within R with the following syntax:
```{r}
df <- data.frame(id = c('a', 'b', 'c'),
                  x = 1:3,
                  y = c(TRUE, TRUE, FALSE),
                  stringsAsFactors = FALSE)
```
Make a data frame that holds the following information about you:

- name
- favorite_color
- year_started_gps

Then use `rbind` to add an entry for the people sitting beside you.
Finally, use `cbind` to add a column with each person's answer to the question, "Are you hungry?"

### Solution to Challenge 1

```{r}
df <- data.frame(name = c('Grace Hopper'),
                 favorite_color = c('red'),
                 year_started_gps = c(2015),
                 stringsAsFactors = FALSE)
df <- rbind(df, list('Marie Curie', 'blue', 2014) )
df <- cbind(df, hungry = c(TRUE,TRUE))
```

---

### Challenge 2 - inclass

Go to file -> new file -> R script, and write an R script
to load in the gapminder dataset. Put it in the `scripts/`
directory and add it to version control.

Run the script using the `source` function, using the file path
as its argument (or by pressing the "source" button in RStudio).

### Solution to Challenge 2

The contents of `script/load-gapminder.R`:
```{r eval = FALSE}
download.file("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv", destfile = "data/gapminder-FiveYearData.csv")
gapminder <- read.csv(file = "data/gapminder-FiveYearData.csv")
```
To run the script and load the data into the `gapminder` variable:
```{r eval = FALSE}
source(file = "scripts/load-gapminder.R")
```


---


### Challenge 3 - in quiz i think

Read the output of `str(gapminder)` again;
this time, use what you've learned about factors, lists and vectors,
as well as the output of functions like `colnames` and `dim`
to explain what everything that `str` prints out for gapminder means.
If there are any parts you can't interpret, discuss with your neighbors!

### Solution to Challenge 3

The object `gapminder` is a data frame with columns

- `country` and `continent` are factors.

- `year` is an integer vector.

- `pop`, `lifeExp`, and `gdpPercap` are numeric vectors.


---

## Lesson 6: Data Subsetting 

---

### Challenge 1 - inclass

Given the following code:

```{r, eval=FALSE}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```

Come up with at least 3 different commands that will produce the following output:

```{r, eval=FALSE, echo=FALSE}
x[2:4]
```

After you find 3 different commands, compare notes with your neighbour. Did you have different strategies?

## Solution to challenge 1

```{r, eval=FALSE}
x[2:4]
```
```{r, eval=FALSE}
x[-c(1,5)]
```
```{r, eval=FALSE}
x[c("b", "c", "d")]
```
```{r, eval=FALSE}
x[c(2,3,4)]
```

---

### Challenge 2 - inclass

Run the following code to define vector `x` as above:

```{r, eval=T}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```

Given this vector `x`, what would you expect the following to do?
```{r, eval=FALSE}
x[-which(names(x) == "g")]
```

Try out this command and see what you get. Did this match your expectation?
Why did we get this result? (Tip: test out each part of the command on it's own - this is a useful debugging strategy)

Which of the following are true:

* A) if there are no `TRUE` values passed to `which`, an empty vector is returned
* B) if there are no `TRUE` values passed to `which`, an error message is shown
* C) `integer()` is an empty vector
* D) making an empty vector negative produces an "everything" vector
* E) `x[]` gives the same result as `x[integer()]`

### Solution to challenge 2

A and C are correct.

The `which` command returns the index of every `TRUE` value in its
input. The `names(x) == "g"` command didn't return any `TRUE` values. Because
there were no `TRUE` values passed to the `which` command, it returned an
empty vector. Negating this vector with the minus sign didn't change its
meaning. Because we used this empty vector to retrieve values from `x`, it
produced an empty numeric vector. It was a `named numeric` empty vector
because the vector type of x is "named numeric" since we assigned names to the
values (try `str(x)` ).

---

### Challenge 3

Given the following code:

```{r, eval=FALSE}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
print(x)
```

Write a subsetting command to return the values in x that are greater than 4 and less than 7.

### Solution to challenge 3

```{r, eval=FALSE}
x_subset <- x[x<7 & x>4]
print(x_subset)
```

---

### Challenge 4 - added to quiz

Given the following code:

```{r, eval=FALSE}
m <- matrix(1:18, nrow=3, ncol=6)
print(m)
```

1. Which of the following commands will extract the values 11 and 14?

A. `m[2,4,2,5]`

B. `m[2:5]`

C. `m[4:5,2]`

D. `m[2,c(4,5)]`

### Solution to challenge 4

D

---


---

### Challenge - 6 in class
Given a linear model:

```{r, eval=FALSE}
mod <- aov(pop ~ lifeExp, data=gapminder)
```

Extract the residual degrees of freedom (hint: `attributes()` will help you)

### Solution to challenge 6

```{r, eval=FALSE}
attributes(mod) ## `df.residual` is one of the names of `mod`
```
```{r, eval=FALSE}
mod$df.residual
```

---



### Challenge 8 - inclass

1. Why does `gapminder[1:20]` return an error? How does it differ from `gapminder[1:20, ]`?


2. Create a new `data.frame` called `gapminder_small` that only contains rows 1 through 9
and 19 through 23. You can do this in one or two steps.

### Solution to challenge 8

1.  `gapminder` is a data.frame so needs to be subsetted on two dimensions. `gapminder[1:20, ]` subsets the data to give the first 20 rows and all columns.

2. 

```{r, eval=FALSE}
gapminder_small <- gapminder[c(1:9, 19:23),]
```

---







